Think deep for reef resilience

### 2025-08-09 paper/paper_draft.qmd
Add small details about DHW methods and make notes for self

Future projections of maximum yearly Degree Heating Weeks (DHWs) for each reef were
estimated from Sea Surface Temperature (SST) data for the GBR. Statistically downscaled 1km
resolution SST data for five Global Climate Models (GCMs) were obtained from the
HighResCoralStress database [@dixon2022; @dixon2023]. These five GCMs were EC-Earth3-Veg,
ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GFDL-CM4 and are a part of the Sixth Phase of the
Coupled Model Intercomparison Project (CMIP6) based on evaluation conducted by Grose et al.
[-@grose2023]. These GCMs were found to rank highly in the recreation of historical
atmospheric temperature and precipitation conditions in the Australasian region. While this
assessment does not consider oceanic data, Grose et al. [-@grose2023] provide a
region-specific selection mechanism which we use as an indicator of general performance of
GCMs for the region. An overview of the selected GCMs are provided in @tbl-gcms. The process
described herein was conducted for all reefs within the GBR Marine Park to create DHW
projections spanning from 2025 to 2075.

@@

DHW climatological baselines were calculated from the SST data following NOAA methodology
[@liu2017; @skirving2020]. For each pixel of SST data, the Monthly Mean SST climatology was
calculated by computing the mean SST value for each calendar month (January through
December) across the 28-year baseline period from 1985 to 2012. This climatological baseline
was then temporally adjusted to center at year 1988.2857, which represents the midpoint of
the reference period and accounts for any long-term trends within the baseline years
themselves. The adjustment ensures that the climatology represents conditions at a specific
temporal reference point rather than being influenced by gradual warming or cooling trends
that may have occurred during the 1985-2012 period. The resulting monthly climatologies
serve as a standardized reference against which current SST conditions can be compared to
identify thermal anomalies and assess coral bleaching risk, following the methodology
established by Skirving et al. [-@skirving2020].

**\[TI: Ben, could you provide more details in the paragraph above please? What resolution
is the data in? 10m\^2?\]**

Degree Heating Weeks (DHWs) were then derived using the "HotSpot SST Anomaly" approach
[@skirving2020]. The "HotSpot SST Anomaly" represents the thermal stress experienced by
coral reefs and is calculated as the difference between daily SST values and the maximum of
a pixel's monthly mean climatology across all twelve months (i.e., the maximum monthly mean
temperature). This approach uses the annual maximum temperature as the baseline because
corals are adapted to withstand their local warmest conditions under normal seasonal cycles.
When temperatures exceed this seasonal maximum, corals experience thermal stress that can
lead to bleaching. Only positive anomalies are considered relevant for coral stress, as
temperatures below the seasonal maximum do not contribute to bleaching risk.

DHW values on a per-pixel basis were then calculated as the 12-week rolling sum of positive
HotSpot SST Anomaly values, expressed in units of °C-weeks. The 12-week accumulation period
reflects the cumulative nature of coral thermal stress. Brief temperature spikes may be
tolerable, but sustained or repeated exposure to elevated temperatures causes progressively
greater physiological stress and bleaching severity. DHW values integrate both the magnitude
and duration of thermal stress: a moderate anomaly (e.g., 1°C) sustained for 12 weeks
produces the same DHW value as a severe anomaly (e.g., 3°C) lasting 4 weeks, though the
biological impacts may differ. DHW estimates for each indicative reef (see @greatba2022)
were then determined by taking the mean DHW value for all pixels within each reef outline.

**\[TI: Ben, I did my best to expand on DHW calculations and detail HotSpot SST anomaly,
could you please check?\]**

::: {#tbl-gcms}
| Global Climate Model source | ECS value (°C) | Interpreted relative severity | Results found in |
|-------------------------|--------------------|---------------------------|--------------------|
| EC-Earth3-Veg | 4.31 | High | **insert repo citation** |
| ACCESS-CM2 | 4.72 | High | **insert repo citation** |
| ACCESS-ESM1-5 | 3.87 | Intermediate | Main text |
| GFDL-CM4 | 2.9 | Low | **insert repo citation** |
| NorESM2-MM | 2.5 | Low | **insert repo citation** |

@@

@@

>>>>

Future projections of maximum yearly Degree Heating Weeks (DHWs) for each reef were
estimated from Sea Surface Temperature (SST) data for the GBR. Statistically downscaled
1km^2^ resolution SST data for five Global Climate Models (GCMs) were obtained from the
HighResCoralStress database [@dixon2022; @dixon2023]. These five GCMs were EC-Earth3-Veg,
ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GFDL-CM4 and are a part of the Sixth Phase of the
Coupled Model Intercomparison Project (CMIP6) based on evaluation conducted by Grose et al.
[-@grose2023]. These GCMs were found to rank highly in the recreation of historical
atmospheric temperature and precipitation conditions in the Australasian region. While this
assessment does not consider oceanic data, Grose et al. [-@grose2023] provide a
region-specific selection mechanism which we use as an indicator of general performance of
GCMs for the region. An overview of the selected GCMs are provided in @tbl-gcms. The process
described herein was conducted for all reefs within the GBR Marine Park to create DHW
projections spanning from 2025 to 2075.

@@

Monthly mean SST climatological baselines, for each 1km^2^ pixel, were calculated from the
daily SST data following NOAA methodology [@liu2017; @skirving2020]. For each pixel of SST
data, the Monthly Mean SST climatology was calculated by computing the mean SST value for
each calendar month (January through December) across the 28-year baseline period from 1985
to 2012. For each month separately, this climatological baseline was then adjusted to center
at year 1988.2857, which attempts to account for any long-term trends within the baseline
years themselves (see Skirving et al. [@skirving2020]). The adjustment ensures that the
climatology represents conditions at a specific temporal reference point rather than being
influenced by gradual warming or cooling trends that may have occurred during the 1985-2012
period. The resulting monthly climatologies serve as a standardized reference against which
current SST conditions can be compared to identify thermal anomalies and assess coral
bleaching risk, following the methodology established by Skirving et al. [-@skirving2020].

Degree Heating Weeks (DHWs) were then derived using the "HotSpot SST Anomaly" approach
[@skirving2020]. The "HotSpot SST Anomaly" represents the thermal stress experienced by
coral reefs and is calculated as the difference between daily SST values and the maximum of
a pixel's monthly mean climatology across all twelve months (i.e., the maximum monthly mean
SST). This approach uses the annual maximum temperature as the baseline because corals are
adapted to withstand their local warmest conditions under normal seasonal cycles. When
temperatures exceed this seasonal maximum, corals experience thermal stress that can lead to
bleaching. Only positive anomalies are considered relevant for coral stress, as temperatures
below the seasonal maximum do not contribute to bleaching risk.

DHW values on a per-pixel basis at 1km^2^ resolution were then calculated as the 12-week
rolling sum of positive HotSpot SST Anomaly values, expressed in units of °C-weeks. The
12-week accumulation period reflects the cumulative nature of coral thermal stress. Brief
temperature spikes may be tolerable, but sustained or repeated exposure to elevated
temperatures causes progressively greater physiological stress and bleaching severity. DHW
values integrate both the magnitude and duration of thermal stress: a moderate anomaly
(e.g., 1°C) sustained for 12 weeks produces the same DHW value as a severe anomaly (e.g.,
3°C) lasting 4 weeks, though the biological impacts may differ. DHW estimates for each
indicative reef (see @greatba2022) were then determined by taking the mean DHW value for all
pixels touching each reef outline.

::: {#tbl-gcms}
| Global Climate Model source | ECS value (°C) | Interpreted relative severity | Results found in |
|------------------------|----------------------|--------------------------|----------------------|
| EC-Earth3-Veg | 4.31 | High | **insert repo citation** |
| ACCESS-CM2 | 4.72 | High | **insert repo citation** |
| ACCESS-ESM1-5 | 3.87 | Intermediate | Main text |
| GFDL-CM4 | 2.9 | Low | **insert repo citation** |
| NorESM2-MM | 2.5 | Low | **insert repo citation** |

@@

\[**BG: Need to add quantification results to text for timeseries differences.**\]

@@

\[**BG: Need to add quantification results to text for regressions.**\]
